---
title: "20250908_vcf_filter_redo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(DiagrammeR)
library(kableExtra)
library(gggenes)
library(cowplot)
```

## Compare the number of unfiltered variants in v2023 and v2025.

To make sure the reproducibility, we re-processed all the samples from raw sequencing data to the final vcf with all intermediate files un-removed.

All samples except one (Plate1E4) have the exact same number of variants. Details see the scripts `pre_01_no_of_variants_stat.R` in the same folder.

In v2023, sample Plate1E4 (Aripiprazole, 50 µM) has only 360 variants while in v2025 in total 645 variants were called. [Log file](/rds-d6/user/rg684/hpc-work/ALE_20220823_sequencing_data_batch1/03.call_variant/20230714_vcf_80X_new_ref/04.snippy_all_fastq/mincov10_C3_F001/Plate1E4/snps.log) shows that bam file had I/O issues (this was not spot before due to the file was only partially corrupt).

All following re-analysis are based on v2025 (and should be the same as v2023 for all samples except Plate1E4).

```{r import, echo=FALSE}
  #v2025 <- readRDS("../03.merge_vcf/20250907_all_variants_redo.rds")
  v2025 <- readRDS("../03.merge_vcf/20250910_all_variants_redo_qual.rds")
  v2025$Evolved <- ifelse(v2025$Compound == "NT5002", "Parental", "Evolved")
```

## Allele Frequency (AF) destribution of all variants. 

Below we show the distribution of allele frequency of all called variants (minimum coverage 10 ` --min-coverage 10`, minimum number of supporting observations 3 to consider a variant `-C 3`, minimum allele frequency 0.01 `-F 0.01`, minimum base quality 30 `-q 30`, minimum mapping quality 60 `-m 60` ) in parental populations (Sample_ID Plate1B9 and Plate1C9) and evolved populations. __Salmon line__ is showing the cutoff of AF 0.05, and __golden line__ is indicating AF 0.5.

```{r AF, echo=FALSE, message=FALSE, error=FALSE, fig.width=7, fig.height=3}
  ggplot(v2025, aes(log(Ratio))) +
    geom_histogram(bins = 100) +
    theme_bw() + labs(x = "Allele Frequency (log-transformed)",
                      y = "Number of variants") +
    facet_wrap(~Evolved, scales = "free") +
    geom_vline(xintercept = log(0.05), color = "salmon", linetype = "dashed")+
    geom_vline(xintercept = log(0.5), color = "gold", linetype = "dashed")
```

## Retrieve the QUAL information from vcf files 

Since the _snps.tab_ files are simplified and reformatted "vcf" files, we went back to _snps.vcf_ file for each sample to retrieve the quality value (QUAL) of variant calling.

QUAL estimates probability of being polymorphic:
- phred 10 means probabilty of polymorphism > 0.9.
- phred 20 means probability of polymorphism > 0.99.
- phred 30 means probability of polymorphism > 0.999.

Below we show the distribution of quality of all variants.

```{r qual_stat, echo=FALSE, message=FALSE, error=FALSE}
  v2025$QUAL_bi <- ifelse(v2025$QUAL == 0, "= 0", 
                          ifelse(v2025$QUAL <= 10, "0-10",
                                 ifelse(v2025$QUAL <= 20, "10-20",
                                        ifelse(v2025$QUAL <= 30, "20-30", ">30"))))  

  stat1 <- v2025 %>% group_by(QUAL_bi, Evolved) %>%
    summarise(No_of_variant = n())
  
  stat1$QUAL_bi <- factor(stat1$QUAL_bi, levels = c("= 0",
                                                    "0-10",
                                                    "10-20",
                                                    "20-30",
                                                    ">30"))
  stat1 <- stat1[order(stat1$QUAL_bi), ]
  kbl(stat1) %>% kable_styling(full_width = F)
```

We will filter out the variants with `QUAL < 1` before plotting the distribution.

```{r qual, echo=FALSE, message=FALSE, error=FALSE, fig.width=7, fig.height=3}
  v2025_filter <- v2025[v2025$QUAL > 1, ]

  ggplot(v2025_filter, aes(log(QUAL))) +
    geom_histogram() +
    theme_bw() + labs(x = "Variant quality (log-transformed)",
                      y = "Number of variants") +
    facet_wrap(~Evolved, scales = "free") +
    geom_vline(xintercept = log(10), color = "salmon", linetype = "dashed") +
    geom_vline(xintercept = log(20), color =  "salmon", linetype = "dashed") +
    geom_vline(xintercept = log(30), color =  "salmon", linetype = "dashed")
    
```

Three dashed lines are indicating Q10, Q20 and Q30 (from left to right). We filter the variants with QUAL < 30.

```{r qual_filter, echo=FALSE, message=FALSE, error=FALSE, fig.width=4, fig.height=3}
  v2025_filter2 <- v2025[v2025$QUAL >= 30, ]
# 
#   kbl(v2025_filter2 %>% group_by(Compound) %>% summarise(No_of_variants = n()),
#                longtable = TRUE)
  filter2_stat <- v2025_filter2 %>% filter(Compound != "NT5002") %>%
    group_by(Compound) %>% summarise(No_of_variants = n())
  
  ggplot(filter2_stat, aes(No_of_variants)) +
    geom_histogram(binwidth = 1) +
    theme_bw() + labs(x = "Number of variant in each population with Q >= 30", y = "Number of evolved popultion")
  
```

As shown in the table, parental strain (`Compound == "NT5002"`) has 12 variants that passed the filter. We examined the AF as well as Ref and Alt depth of these variants.

```{r parental, echo=FALSE, message=FALSE, error=FALSE}
  parental <- v2025_filter2[v2025_filter2$Compound == "NT5002", ]

  kbl(parental[, c("Sample_ID", "POS", "QUAL", "Ref_depth", "Alt_depth",
                            "EFFECT", "PRODUCT")]) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

As we can see, most of the variants here are detected in both sequenced samples (two technical replicates), except for two positions. 

```{r parental_stat, echo=FALSE, message=FALSE, error=FALSE}
  kbl(parental %>% group_by(POS) %>% 
                 summarise(Count = n(), Avg_AF = mean(Ratio))) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

Previously, we defined:

>For the variant that appeared in both parental short-reads sequenced samples and with average AF higher than 90%, we identify those >as “parental variants” (including variants at the positions: 1166763, 1681625, 1702756, 3356587) that are highly possible false >positive due to mis-assembly. 

After manually check the mapping results (`snps.report.txt` files), all positions but _1914986_ have solid mapping results, so we will update the "parental variants" list: 

- previous ones 1166763, 1681625, 1702756, 3356587 
- potentially: plus newly added after releasing filtering & manual check: 1931428, 3256546 (Note: these two only detected in one of the parental samples)

Below we show the prevalence of these 6 variants in all samples:

```{r parental_variants_prevalence, echo=FALSE, message=FALSE, error=FALSE, fig.width=6, fig.height=3}
    parental_lst <- c("1166763", "1681625", "1702756", 
                    "1931428", "3256546", "3356587")
   parental_var <- v2025_filter2[v2025_filter2$POS %in% parental_lst, ]
   
    ggplot(parental_var, aes(as.factor(POS), Ratio)) + geom_boxplot(color = "gray47") +
      geom_jitter(color = "gray28") + theme_bw() + labs(x = "Position", y = "AF")
    
    kbl(parental_var %>% group_by(POS) %>% summarise(Prevalence = n())) %>%
      kable_styling(bootstrap_options = "striped", full_width = F, position = "float_right")
```

__Update__: we add/label two more variants as "parental variants" here.

## Filter out parental variants from evolved populations

We then filter out the parental variants from evolved populations. 

```{r filter_parental, echo=FALSE, message=FALSE, error=FALSE, fig.width=7, fig.height=3}
  evolved_var <- v2025_filter2[v2025_filter2$Compound != "NT5002", ]
  ## 1353 30
  evolved_var <- evolved_var[!(evolved_var$POS %in% parental_lst), ]
  ## 128 30
  
  evolved_var_stat <- evolved_var %>% group_by(Sample_ID, Compound, Concentration) %>%
    summarise(No_of_variants = n())
 
  ggplot(evolved_var_stat, aes(Compound, No_of_variants)) +
    #geom_boxplot() +
    geom_jitter(aes(color = Concentration)) +
    theme_bw() + labs(x = "Evolved populations",
                      y = "Number of variants") +
    scale_color_manual(values = c("gray", "deepskyblue", "deepskyblue4")) +
    theme(legend.position = "top",
          axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

To __conclude__: the _re-filter-1_ we tried here

```{r filter_1, echo=FALSE, message=FALSE, error=FALSE, fig.width=6, fig.height=2}
  mermaid("
  graph LR;
      A[All variants] --> B[QUAL >= 30];
      B --> C[Filter out 6 parental variants]
  ")
```

Using the new filter (QUAL > 30, `r nrow(v2025_filter2)` variants remained; `r nrow(evolved_var)` variants remained after removing parental 6 variants), we dramatically decreased the false positive; but meanwhile the false negative is hugely increased.
See details about this in next session (__Figure 2d-e__).

## Xanthan gum related mutations 

We here show the mapping of genetic variants detected in populations evolved with
xanthan gum, as we reported in __Figure 2d__ in current version manuscript.

Using previous filtering:

```{r filter_0, echo=FALSE, message=FALSE, error=FALSE, fig.width=5, fig.height=2}
  mermaid("
  graph LR;
      A[All variants] --> B[Allele frequency > 0.05];
      B --> C[Filter out 4 parental variants] 
  ")
```

We showed that multiple samples with the variants detected at the same position.

```{r figure2d, echo=FALSE, message=FALSE, error=FALSE, fig.width=4, fig.height=3}
    pos_gene <- read.table("../../00.data/vcf_20231025_position_gene_map.txt",
                           header = F, sep = "\t")
    colnames(pos_gene) <- c("POS", "Type", "Gene")
    map <- read.table("../../00.data/vcf_20231024_effect_map.txt", 
                      header = T, sep = "\t")
    map$Effect_type <- as.factor(map$Effect_type)
    gff <-read.delim("../../00.data/annot_RAST_simp.gff", header = F, sep = "\t")
    colnames(gff) <- c("Start", "End", "Strand", "Gene_ID", "Annotation")
    
    v2025_filter3 <- v2025[v2025$Ratio > 0.05, ]
    
    f2d <- merge(v2025_filter3, pos_gene)
    
    #### Panel D - AcrR 
    this_gene_ID <- c("peg.1416", "peg.1417")
    
    this_gene <- f2d[f2d$Gene %in% this_gene_ID, ]
    this_gene <- merge(this_gene, map)
    
    ## Get the coordinate of this gene
    this_gff <- gff[gff$Gene_ID %in% this_gene$Gene, ]
    this_start <- min(this_gff$End)
    this_end <- max(this_gff$End)
    
    p1 <- ggplot(this_gene, aes(POS, Ratio)) +
      geom_point(aes(color = Compound, shape = Effect_type, size = Concentration),
                 alpha = 0.9) + 
      xlim(this_start, this_end) +
      labs(x = "", y = "AF") +
      scale_size_manual(values = c("0" = 1,
                                   "25" = 1.8, 
                                   "50" = 2.4), guide = F) +
      scale_color_manual(values = c("Xanthan_gum" = "slateblue3",
                                    "Others" = "gray")) +
      scale_shape_manual(limits = map$Effect_type, values = map$Shape, guide = F) +
      theme_bw() +
      theme(legend.position = "top") +
      guides(color = guide_legend(nrow = 2)) +
      theme(legend.background = element_blank())
    
    colnames(this_gff) <- c("start", "end", "strand", "gene", "protein")
    this_gff$orientation <- ifelse(this_gff$strand == "+", 1, 0)
    this_gff$molecule <- " "
    
    p2 <- ggplot(this_gff, aes(xmin = start, xmax = end,
                               y = molecule, fill = protein, 
                               forward = orientation)) +
      xlim(this_start, this_end) +
      geom_gene_arrow() + labs(y = "") +
      scale_fill_brewer(palette = "Purples") +
      theme_genes() +
      theme(legend.position = "none",
            panel.background = element_blank(),
            plot.background = element_blank()) 
    
    plot_grid(p1, p2, nrow = 2, align = "v",  axis = "l", 
                     rel_heights = c(4, 1))
```

The detailed information about these variants are shown below:

```{r figure2d_tab, echo=FALSE, message=FALSE, error=FALSE}
  kbl(this_gene[, c("Sample_ID", "POS", "QUAL", "Ref_depth", "Alt_depth", "Ratio",
                    "Compound", "Concentration",
                            "EFFECT", "PRODUCT")]) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```