---
title: "20251021_vcf_filter_strict"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(DiagrammeR)
library(kableExtra)
library(gggenes)
library(cowplot)
source("../../02.scripts_variants/settings.R")
```

## Compare the number of unfiltered / raw variants in v2023 and v2025

To make sure the reproducibility, we re-processed all the samples from raw sequencing data to the final vcf with all intermediate files un-removed.

All samples except one (Plate1E4) have the exact same number of variants. Details see the scripts `pre_01_no_of_variants_stat.R` in the same folder.

In v2023, sample Plate1E4 (Aripiprazole, 50 µM) has only 360 variants while in v2025 in total 645 variants were called. [Log file](/rds-d6/user/rg684/hpc-work/ALE_20220823_sequencing_data_batch1/03.call_variant/20230714_vcf_80X_new_ref/04.snippy_all_fastq/mincov10_C3_F001/Plate1E4/snps.log) shows that bam file had I/O issues (this was not spot before due to the file was only partially corrupt).

All following re-analysis are based on v2025 (and should be the same as v2023 for all samples except Plate1E4).

```{r import, echo=FALSE}
  #v2025 <- readRDS("../03.merge_vcf/20250907_all_variants_redo.rds")
  v2025 <- readRDS("../03.merge_vcf/20250910_all_variants_redo_qual.rds")
  v2025$Evolved <- ifelse(v2025$Compound == "NT5002", "Parental", "Evolved")
```

Later on, after filtering, no variant was kept for sample Plate1E4.

## Add the QUAL information from vcf files 

Since the _snps.tab_ files (previously used for Figure 2) are simplified and reformatted "vcf" files, we went back to _snps.vcf_ file for each sample to retrieve the quality value (QUAL) of variant calling.

QUAL estimates probability of being polymorphic:  
- phred 10 means probabilty of polymorphism > 0.9.  
- phred 20 means probability of polymorphism > 0.99.  
- phred 30 means probability of polymorphism > 0.999.  

## Filter raw vcf files 

Below we show a summarized workflow of the whole process. Details of number and distribution can be find in [20250908_vcf_filter.html](20250908_vcf_filter.html).

```{r filter_by_alt_depth, echo=FALSE, message=FALSE, error=FALSE, fig.width=12, fig.height=2}
  mermaid("
  graph LR;
      A[All variants] --> |486155|B[Alt_depth >= 5];
      B --> |23885|C[Allele frequency > 0.05];
      C --> |2115|D[Identify parental variants N = 15];
      D --> E[Remove parental variants];
      E --> |686|F[Evolved variants];
      F --> |199|G[Keep variants found in <= 3 compounds]
  ")
```

Below we summarize the distribution of quality (QUAL) of filtered variants.

```{r filtered_qual,  warning=FALSE, echo=FALSE, message=FALSE}
  filtered_vcf <- readRDS("20251021_filtered_vcf_199.rds")
  kbl(filtered_vcf %>% group_by(QUAL_bi) %>% summarise(No. = n())) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

As we can see, if we want to add the QUAL filtering (QUAL > 30, Filter option 1 in [20250908_vcf_filter.html](20250908_vcf_filter.html)), only 8 variants at 6 different
positions will be kept. 

We therefore, instead of filtering by quality, use IGV to manually check each variants shown in Figure 2 panels related to Xanthan gum and Loperamide.

## Update Figure 2 with new filtering

We then re-plotted figure 2 with the new filtering strategy. Since only `r length((filtered_vcf$POS))` variants at `r length(unique(filtered_vcf$POS))` are kept, we would not need two panels (previous Figure 2a and 2b) to show the landscape of variants. 

Since the enrichment of SusC/D was filtered out, we will remove the previous panel c.

The new figure layout looks:  
- __a__ variant landscape over the genome that pass the strict filtering, with Xanthan gum and Loperamide related gene region highlighted   
- __b__ Xanthan gum evolved strains' variants associated to _acrR_ gene  
- __c__ Loperamide evolved strains' variants associated to _trkH_ and _trkA_ genes.


```{r figure_2abc, warning=FALSE, echo=FALSE, message=FALSE}
  all <- filtered_vcf
  t1 <- all %>% group_by(POS) %>% summarise(Prevalence =
                                              length(unique(Sample_ID)),
                                            Average_AF = mean(Ratio))
  t1_des <- unique(all[, c("POS", "FTYPE", "Effect")])
  t1 <- merge(t1, t1_des)
  
  map <- read.table("../../00.data/vcf_20251021_effect_map.txt", 
                    header = T, sep = "\t")
  map$Effect_type <- as.factor(map$Effect_type)
  #map$Shape <- as.factor(map$Shape)
  
  t1 <- merge(t1, map)
  
  p1_1 <- ggplot(t1, aes(POS, Prevalence)) +
    annotate("rect", ymin=-Inf, ymax=Inf, xmin=1771711, xmax=1772529, 
             alpha=0.8, fill="slateblue3") + 
    annotate("rect", ymin=-Inf, ymax=Inf, xmin=2104963, xmax=2107776, 
             alpha=0.8, fill="#98dd94") + 
    geom_point(aes(color = FTYPE, size = Average_AF, shape = Effect_type),
               alpha = 0.8) + 
    scale_shape_manual(limits = map$Effect_type,
                       values = map$Shape) +
    scale_color_manual(values = c("gold", "gray47")) +
    xlim(c(0, 4.685e+06)) + xlab("") +
    scale_size(range = c(0.25, 5), limits = c(0.05, 1)) +
    theme_minimal_hgrid(color = "gray88", font_size = 10) +
    guides(colour = guide_legend(order = 2, nrow = 1), 
           size = guide_legend(order = 3, nrow = 1),
           shape = guide_legend(order = 1, nrow = 1)) +
    main_theme + theme(legend.position = "top")
  
  legend <- get_plot_component(p1_1, 'guide-box-top', return_all = TRUE)
  p1_1 <- p1_1 + theme(legend.position = "none")
  
  p1_2 <- ggplot(all, aes(POS)) + 
    geom_density(aes(color = FTYPE, y = ..scaled..)) +
    scale_color_manual(values = c("gold", "gray47")) +
    scale_y_continuous(position = "right") +
    xlab("") + ylab("Density of variants") +
    xlim(c(0, 4.685e+06)) +
    theme_minimal_hgrid(color = "gray88", font_size = 10) +
    theme(legend.position = "none",
          axis.text.x=element_blank()) +
    main_theme
  
  aligned_plots <- align_plots(p1_2, p1_1, align="hv", axis="tblr")
  p_a <- ggdraw(aligned_plots[[2]]) + draw_plot(aligned_plots[[1]])
  p_a <- plot_grid(legend, p_a, nrow = 2, rel_heights = c(0.2, 1))
```


```{r figure_2de_filter2, warning=FALSE, echo=FALSE, message=FALSE, fig.height=6, fig.width=7}
  library(gggenes) 
  
  gff <-read.delim("../../00.data/annot_RAST_simp.gff", header = F, sep = "\t")
  colnames(gff) <- c("Start", "End", "Strand", "Gene_ID", "Annotation")
  
  pos_gene <- read.table("../../00.data/vcf_20231025_position_gene_map.txt",
                         header = F, sep = "\t")
  colnames(pos_gene) <- c("POS", "Type", "Gene")
  all <- merge(all, pos_gene)
  
  #### Panel D - AcrR 
  this_gene_ID <- c("peg.1416", "peg.1417")
  
  this_gene <- all[all$Gene %in% this_gene_ID, ]
  this_gene <- merge(this_gene, map)
  
  ## Get the coordinate of this gene
  this_gff <- gff[gff$Gene_ID %in% this_gene$Gene, ]
  this_start <- min(this_gff$End)
  this_end <- max(this_gff$End)
  
  p1 <- ggplot(this_gene, aes(POS, Ratio)) +
    geom_point(aes(color = Compound, shape = Effect_type, size = Concentration),
               alpha = 0.9) + 
    xlim(this_start, this_end) +
    labs(x = "", y = "AF") +
    scale_size_manual(values = c("0" = 1,
                                 "25" = 1.8, 
                                 "50" = 2.4), guide = F) +
    scale_color_manual(values = c("Xanthan_gum" = "slateblue3",
                                  "Others" = "gray")) +
    scale_shape_manual(limits = map$Effect_type, values = map$Shape, guide = F) +
    theme(legend.position = "top") +
    guides(color = guide_legend(nrow = 2)) +
    main_theme +
    theme(legend.background = element_blank())
  
  colnames(this_gff) <- c("start", "end", "strand", "gene", "protein")
  this_gff$orientation <- ifelse(this_gff$strand == "+", 1, 0)
  this_gff$molecule <- " "
  
  p2 <- ggplot(this_gff, aes(xmin = start, xmax = end,
                             y = molecule, fill = protein, 
                             forward = orientation)) +
    xlim(this_start, this_end) +
    geom_gene_arrow() + labs(y = "") +
    scale_fill_brewer(palette = "Purples") +
    theme_genes() +
    theme(legend.position = "none",
          panel.background = element_blank(),
          plot.background = element_blank()) 
  
  p_b <- plot_grid(p1, p2, nrow = 2, align = "v",  axis = "l", 
                   rel_heights = c(4, 1))
  
  ###############################################################################
  #### Panel E
  this_gene_ID <- c( "peg.1703", "peg.1704")
  
  this_gene <- all[all$Gene %in% this_gene_ID, ]
  this_gene <- merge(this_gene, map)
  
  ## Get the coordinate of this gene
  this_gff <- gff[gff$Gene_ID %in% this_gene$Gene, ]
  this_start <- min(this_gff$Start)
  this_end <- max(this_gff$End)
  
  p1 <- ggplot(this_gene, aes(POS, Ratio)) +
    geom_point(aes(color = Compound, shape = Effect_type, size = Concentration),
               alpha = 0.9) + 
    xlim(this_start, this_end) +
    labs(x = "", y = "AF") +
    scale_size_manual(values = c("0" = 1,
                                 "25" = 1.8, 
                                 "50" = 2.4), guide = F) +
    scale_color_manual(values = c("Loperamide" = "#98dd94",
                                  "Others" = "gray")) +
    scale_shape_manual(limits = map$Effect_type, values = map$Shape, guide = F) +
    theme(legend.position = "top") +
    guides(color = guide_legend(nrow = 2)) +
    main_theme +
    theme(legend.background = element_blank())
  
  colnames(this_gff) <- c("start", "end", "strand", "gene", "protein")
  this_gff$orientation <- ifelse(this_gff$strand == "+", 1, 0)
  this_gff$molecule <- " "
  
  p2 <- ggplot(this_gff, aes(xmin = start, xmax = end,
                             y = molecule, fill = protein, 
                             forward = orientation)) +
    xlim(this_start, this_end) +
    geom_gene_arrow() + labs(y = "") +
    scale_fill_manual(values = c("#98dd94", "#72a66f")) +
    theme_genes() +
    theme(legend.position = "none",
          panel.background = element_blank(),
          plot.background = element_blank()) 
  
  p_c <- plot_grid(p1, p2, nrow = 2, align = "v",  axis = "l", 
                   rel_heights = c(4, 1))
  
  ###############################################################################
  ####
  p_bc <- plot_grid(p_b, p_c, nrow = 1, align = "h", axis = "b", 
                     labels = c("b", "c"))
  
  fig2 <- plot_grid(p_a, p_bc, nrow = 2, align="v", axis="lr", labels = c("a", ""),
                    rel_heights = c(1, 1.2))
  
  fig2
```

