---
title: "20250814_R1_vcf_filter"
output: html_document
date: "2025-08-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("settings.R")
```

## VCF filtering - Figure 2a,b

In previous version, we have:

>"In total, 10,041 variants, with an allele frequency (AF) higher than 0.05, were
>identified at 6,097 different positions across all evolved populations (including
>the solvent-control conditions), with a median of 28 variants (Figures 2a; EV4a, b)."

According to our discussion last time (Simone, Sonja, Rui; 2025.08.06); we are 
filtering variants that were appeared in both control and xenobiotic groups; as 
those are highly likely to be _pre-exist_ variants.

We have mentioned/discussed this in the past that __variants__ we identified includes:
1) newly emerged mutations
2) pre-exist genetic variants emerged / sequenced due to increased ratio in population

One reason we have the above 2) is our parental strain __was not__ single colony.

Now, with new filtering:

1) Ref + Alt depth >= 20;
2) Alt depth >= 5; __this filtered out the most of the variants__
3) Positions with prevalence < 24 (i.e. max. appeared in 3 diff. xenobiotics)
`this will be refined`

```{r vcf_filter, message=FALSE, warning=FALSE, echo=FALSE, fig.width=9, fig.height=3}
  all <- readRDS("../04.results/20231023_all_005.rds")
  all <- all[all$Exp == "ALE1", ]
  
  all$Depth <- all$Ref_depth + all$Alt_depth
  all <- all[all$Depth >= 20, ]
  # 9978 
  all <- all[all$Alt_depth >= 5, ]
  #dim(all)
  # 1776
  
  t1 <- all %>% group_by(POS) %>% summarise(Prevalence =
                                              length(unique(Sample_ID)),
                                            Average_AF = mean(Ratio))
  t1_des <- unique(all[, c("POS", "FTYPE", "Effect")])
  t1 <- merge(t1, t1_des)
  
  t1_filter <- t1[t1$Prevalence < 24, ]
  all_filter <- all[all$POS %in% t1_filter$POS, ]
  # 868 
  
  map <- read.table("../00.data/vcf_20231024_effect_map.txt", 
                  header = T, sep = "\t")
  map$Effect_type <- as.factor(map$Effect_type)

  t1_filter <- merge(t1_filter, map)
  
  p1_1 <- ggplot(t1_filter, aes(POS, Prevalence)) +
  annotate("rect", ymin=-Inf, ymax=Inf, xmin=1910551, xmax=1944637, 
           alpha=0.3, fill="skyblue") + 
  geom_point(aes(color = FTYPE, size = Average_AF, shape = Effect_type),
             alpha = 0.7) + 
  scale_shape_manual(limits = map$Effect_type,
                     values = map$Shape) +
  scale_color_manual(values = c("gold", "gray47")) +
  xlim(c(0, 4.685e+06)) + xlab("") +
  scale_size(range = c(0.25, 5), limits = c(0.05, 1)) +
  theme_minimal_hgrid(color = "gray88", font_size = 10) +
  guides(colour = guide_legend(order = 2, nrow = 1), 
         size = guide_legend(order = 3, nrow = 1),
         shape = guide_legend(order = 1, nrow = 1)) +
  main_theme + theme(legend.position = "top")

  legend <- get_plot_component(p1_1, 'guide-box-top', return_all = TRUE)
  p1_1 <- p1_1 + theme(legend.position = "none")

  p1_2 <- ggplot(all_filter, aes(POS)) + 
    geom_density(aes(color = FTYPE, y = ..scaled..)) +
    scale_color_manual(values = c("gold", "gray47")) +
    scale_y_continuous(position = "right") +
    xlab("") + ylab("Density of variants") +
    xlim(c(0, 4.685e+06)) +
    theme_minimal_hgrid(color = "gray88", font_size = 10) +
    theme(legend.position = "none",
          axis.text.x=element_blank()) +
    main_theme
  
  aligned_plots <- align_plots(p1_2, p1_1, align="hv", axis="tblr")
  p_a <- ggdraw(aligned_plots[[2]]) + draw_plot(aligned_plots[[1]])
  p_a <- plot_grid(legend, p_a, nrow = 2, rel_heights = c(0.2, 1))
  
  p_a
```

After filtering, `r nrow(all_filter)` variants in `r length(unique(all_filter$POS))`
positions were remained. As we can see, most of them appeared in one sample.

```{r prevalence, echo = F, message=FALSE, warning=FALSE, fig.width= 4, fig.height=3}
  ggplot(t1_filter, aes(Prevalence)) + geom_histogram(bins = 30) + theme_bw()
```

We here print out the positions with Prevalence > 10.

```{r prevalent, echo = F, message=FALSE, warning=FALSE}
  t1_filter_prevalent <- t1_filter[t1_filter$Prevalence >= 10, ]
  rmarkdown::paged_table(t1_filter_prevalent)
  
  all_prevalent <- all_filter[all_filter$POS %in% t1_filter_prevalent$POS, ]
  all_prevalent <- all_prevalent[order(all_prevalent$POS), ]
  #rmarkdown::paged_table(all_prevalent)
```

## VCF filtering - Figure 2a,b

Plot the prevalence of variants in SusC/D region

```{r suscd, echo = F, message=FALSE, warning=FALSE, fig.width= 4, fig.height=3}
library(gggenes) 
all <- readRDS("../04.results/20231023_all_005.rds")
all <- all[all$Exp == "ALE1", ]

gff <-read.delim("../00.data/annot_RAST_simp.gff", header = F, sep = "\t")
colnames(gff) <- c("Start", "End", "Strand", "Gene_ID", "Annotation")

#t3 <- t1[t1$Prevalence < 10, ]
#all_filter_3 <- all[all$POS %in% t3$POS, ]
#region <- all_filter[(all_filter$POS > 1910551) & (all_filter$POS < 1944637), ]

region <- all[(all$POS > 1910551) & (all$POS < 1944637), ]
region <- merge(region, map)

region_stat <- region %>% group_by(POS) %>% summarise(Prevalence =
                                              length(unique(Sample_ID)),
                                            Average_AF = mean(Ratio))
ggplot(region_stat, aes(Prevalence)) + geom_histogram(bins = 30) + theme_bw()
```

As we can see,  most of the variants appeared in one sample, we will plot these only below:

```{r singleton, echo = F, message=FALSE, warning=FALSE, fig.width= 9, fig.height=3}
gene_lst <- paste0("peg.", seq(1535, 1563))
region_gff <- gff[gff$Gene_ID %in% gene_lst, ]

## Get the coordinate of this gene
this_start <- min(region_gff$Start)
this_end <- max(region_gff$End)

colnames(region_gff) <- c("start", "end", "strand", "gene", "protein")

region_gff$orientation <- ifelse(region_gff$strand == "+", 1, 0)
region_gff$molecule <- ""

p_c1 <- ggplot(region_gff, aes(xmin = start, xmax = end,
                             y = molecule, fill = protein, 
                             forward = orientation)) +
  geom_gene_arrow() + labs(y = "") +
  xlim(this_start, this_end) +
  scale_fill_brewer(palette = "Set3") +
  theme_genes() +
  theme(legend.position = "none")

region_stat_single <- region_stat[region_stat$Prevalence == 1, ]
region_filter <- region[region$POS %in% region_stat_single$POS, ]
region_filter <- region_filter[!(region_filter$Sample_ID %in% c("Plate1B9",
                                           "Plate1C9")), ]



##AF of each variants from each compounds
p_c2 <- ggplot(region_filter, aes(POS, Ratio)) +
  geom_point(aes(color = Concentration, shape = Effect_type), 
             size = 3, alpha = 0.8) +
  xlim(this_start, this_end) +
  scale_shape_manual(limits = map$Effect_type,
                     values = map$Shape, guide = F) +
  scale_color_manual(values = c("0" = "gray",
                                "25" = "#a0cbe8", 
                                "50" = "#1170aa")) +
  #scale_size(range = c(0.25, 5), guide = F) +
  labs(x = "", y = "Allele Frequence (AF)") +
  theme_bw() +
  theme(legend.position = "top", 
        legend.justification = c("left", "center")) +
  #guides(shape = guide_legend(nrow = 1)) +
  theme(legend.background = element_blank()) 

p_c <- plot_grid(p_c2, p_c1, nrow = 2, rel_heights = c(4, 1),
          align = "v", axis = "l")
p_c
```

## Update 2025.08.27